WSClean: a new implementation of a fast, generic wide-field imager

### 2014-01-21 articles/2014/Article-WSClean/wsclean.tex
Minor changes
Adding images
Work on article

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's w-projection implementation. To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms \citep{widefield-imaging-ska-cornwell} combined with snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''. This works improves the field of wide-field imaging in the following ways:
\begin{itemize}
 \item The performance of a w-stacking implementation is compared to a w-projection implementation.
 \item The advantages of using the w-snapshots algorithm are analysed.
 \item Details are described of how to efficiently implement the w-stacking and w-snapshots algorithms.
 \item A new implementation is released, which we believe to be the fastest generic widefield imager available for many imaging scenarios.
\end{itemize}

@@

\subsection{Discretization of w} \label{sec:gridding-w}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of w-layers that need to be processed. For this, we can use a similar rule to \eqref{eq:when-2d-is-valid}, and make sure that for two subsequent discretized $w$-values, $w_A$ and $w_B$, the phase difference is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggest that a uniform discretization in $w$ is optimal TODO in contrast to some article? -- could be explained by having more samples at smaller $w$s. From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation}
 N_\textrm{wlayers} \gg 2\pi \left(w_e - w_0\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can differ a lot. The value of $w_e - w_0$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_e - w_0$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$. [TODO table of arrays with $w$ values at zenith/more angles/more freq?] For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$. This implies that, in the case of the MWA, tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in section TODO.

TODO: Algorithm; plain w-stacking and further alterations such as w-snapshots and full-sky imaging.
\subsection{Gridding}
TODO

@@

\section{The WSClean imager implementation} \label{sec:implementation}
We have implemented the $w$-stacking algorithm in the C++ language. Several steps are multi-threaded in this implementation: reading and writing, gridding from and to different $w$-layers, performing the Fast Fourier transforms and peak-finding and image subtraction during Cleaning. In addition, intrinsics were used for the latter. This speeds up a minor Cleaning iteration,

@@

>>>>

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's w-projection implementation. To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms \citep{widefield-imaging-ska-cornwell} combined with snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''. This work improves the field of wide-field imaging in the following ways:
\begin{itemize}
 \item The performance of a w-stacking implementation is compared to a w-projection implementation.
 \item The advantages of using the w-snapshots algorithm are analysed.
 \item Details are described of how to efficiently implement the w-stacking and w-snapshots algorithms.
 \item A new implementation is released, which we believe to be the fastest generic widefield imager available for many imaging scenarios.
\end{itemize}

@@

\subsection{Discretization of w} \label{sec:gridding-w}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of w-layers that need to be processed. For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggest that a uniform discretization in $w$ is optimal TODO in contrast to some article? -- could be explained by having more samples at smaller $w$s. From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation}
 N_\textrm{wlayers} \gg 2\pi \left(w_e - w_0\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can differ a lot. The value of $w_e - w_0$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_e - w_0$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$. [TODO table of arrays with $w$ values at zenith/more angles/more freq?] For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$. This implies that, in the case of the MWA, tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in section TODO.

TODO: Algorithm; plain w-stacking and further alterations such as w-snapshots and full-sky imaging.

@@

\section{The WSClean imager implementation} \label{sec:implementation}
We have implemented the $w$-stacking algorithm in the C++ language. Several steps are multi-threaded in this implementation: reading and writing, gridding from and to different $w$-layers, performing the Fast Fourier transforms and peak-finding and image subtraction during Cleaning. In addition, intrinsics were used for the latter. This speeds up a minor Cleaning iterations,

@@

\subsection{Gridding}
During gridding of the visibilities in $uv$-space, it is useful to apply a gridding convolution kernel. A common kernel function is a windowed sinc function. A low-pass filter decreases the flux of sources outside the field of view, and thus helps to attenuate aliased ghost sources and sidelobes (see \citealt{post-correlation-filtering}) that arise because of the assumed periodicity during the inverse FFT. By supersampling, a convolution kernel also makes it possible to place samples more accurately at their $uv$ position, thereby lowering decorrelation.

The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding (TODO cite). CASA convolves samples with a PSWF of 7 pixels total width during gridding. When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate\footnote{CASA uses pre-calculation tables for its fixed-size window function}. WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and its properties are similar to a PSWF. The type of window function has no effect on the gridding performance, but gridding has a quadratic time dependency on the size of the kernel.

\begin{table}
 \caption{Parameter values during benchmarks (unless the effect of the parameter was tested).}
 \begin{tabular}{lr}
   \textbf{Parameter} & \textbf{Value} \\
   Image size             & 3072 \\
   Number of visibilities & $10^9$
 \end{tabular}
\end{table}

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-zenith-angle/za}
\caption{...}
\label{fig:timing-zenith-angle}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-nsamples/nsamples}
\caption{...}
\label{fig:timing-nsamples}
\end{center}
\end{figure}
